---
title: "ainsel weather"
author: "Ainsel Levitskaia-Collins, HL2710"
date: "2025-11-18"
output: github_document
---

```{r include = FALSE}
library(tidyverse)
```

### Initial Setup

Importing Data:

```{r message = FALSE, warning = FALSE}
shooting_h <- read_csv("./Data Folder/Shooting_Historic.csv") %>% 
  janitor::clean_names()

shooting_2025 <- read_csv("./Data Folder/Shooting_2025.csv") %>% 
  janitor::clean_names() %>% 
  mutate(statistical_murder_flag = case_match(statistical_murder_flag,
                                             "N" ~ FALSE,
                                             "Y" ~ TRUE))

weather <- read_csv("./Data Folder/Weather.csv") %>% 
  janitor::clean_names()

holidays <- readxl::read_excel("./Data Folder/Holidays.xlsx")
```

Merge & tidy shooting data:

```{r}
shooting <- bind_rows(shooting_2025, shooting_h) %>% 
  separate_wider_delim(occur_date, delim = "/", names_sep = ".") %>% 
  rename(
    month = occur_date.1,
    day = occur_date.2,
    year = occur_date.3,
    time = occur_time
  ) %>% 
  mutate(
    month = case_match(month,
      "01" ~ "January",
      "02" ~ "February",
      "03" ~ "March",
      "04" ~ "April",
      "05" ~ "May",
      "06" ~ "June",
      "07" ~ "July",
      "08" ~ "August",
      "09" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December"),
    day = as.numeric(day),
    year = as.numeric(year)) %>% 
  distinct(incident_key, .keep_all = TRUE) %>% 
  filter(boro == "MANHATTAN")
```



### Weather Analysis

Tidy weather data:

```{r}
weather <- weather %>% 
  separate_wider_delim(date, delim = "-", names_sep = "_") %>% 
  rename(
    year = date_1,
    month = date_2,
    day = date_3
  ) %>% 
  mutate(
    month = case_match(month,
      "01" ~ "January",
      "02" ~ "February",
      "03" ~ "March",
      "04" ~ "April",
      "05" ~ "May",
      "06" ~ "June",
      "07" ~ "July",
      "08" ~ "August",
      "09" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December"),
    day = as.numeric(day),
    year = as.numeric(year)) %>% 
  filter(year >= 2006) %>% 
  select(name, month, day, year, prcp, prcp_attributes, tmax, tmin, latitude, longitude)
```

Merge with shooting data:

```{r}
shooting_weather <- full_join(shooting, weather, by = c("month", "day", "year")) %>% 
  mutate(
    shooting = incident_key > 0,
    shooting = case_match(shooting,
                          TRUE ~ "Incident",
                          NA ~ "No Incident")
  )
```

Plot of max/min temperature and whether there was a shooting:

```{r message = FALSE}
plot_temp_shooting <- shooting_weather %>% 
  arrange(desc(shooting)) %>% 
  ggplot(aes(x = tmin, y = tmax, color = shooting)) +
  geom_point() +
  scale_color_manual(values = c("violetred4", "lightsteelblue3")) +
  theme_minimal() +
  xlab("Minimum daily temperature (tenths of degrees C)") +
  ylab("Maximum daily temperature (tenths of degrees C)") +
  ggtitle("Shooting incidence compared against daily temperature")

plot_temp_shooting
```

No visible correlation between temperature and shooting incidence.

Plot of precipitation and whether there was a shooting:

```{r message = FALSE}
plot_prcp_low_shooting <- shooting_weather %>% 
  filter(prcp < 400) %>% 
  ggplot(aes(x = prcp, color = shooting, fill = shooting)) + 
  geom_histogram() +
  scale_color_manual(values = c("violetred4", "lightsteelblue3")) +
  scale_fill_manual(values = c("violetred4", "lightsteelblue3")) +
  theme_minimal() +
  xlab("Precipitation (tenths of mm)") +
  ylab("") +
  ggtitle("Shooting count in low precipitation")

plot_prcp_low_shooting

plot_prcp_high_shooting <- shooting_weather %>% 
  filter(prcp > 400) %>% 
  ggplot(aes(x = prcp, color = shooting, fill = shooting)) + 
  geom_histogram() +
  scale_color_manual(values = c("violetred4", "lightsteelblue3")) +
  scale_fill_manual(values = c("violetred4", "lightsteelblue3")) +
  theme_minimal() +
  ylab("") +
  xlab("Precipitation (tenths of mm)") +
  ggtitle("Shooting count in high precipitation")

plot_prcp_high_shooting
```

There appears to be a higher quantity of shootings during lower-precipitation days, however it's difficult to tell whether this is correlative or not. 

- [ ] linear regression? to see whether this is a significant relationship

### Holiday Analysis

Tidy holiday data:

```{r}
holidays <- holidays %>% 
  pivot_longer(`2025`:`2006`, names_to = "year", values_to = "date") %>% 
  separate_wider_delim(date, delim = "/", names_sep = "_") %>% 
  rename(
    month = date_1,
    day = date_2
  ) %>% 
  mutate(
    month = case_match(month,
      "1" ~ "January",
      "2" ~ "February",
      "3" ~ "March",
      "4" ~ "April",
      "5" ~ "May",
      "6" ~ "June",
      "7" ~ "July",
      "8" ~ "August",
      "9" ~ "September",
      "10" ~ "October",
      "11" ~ "November",
      "12" ~ "December"),
    day = as.numeric(day),
    year = as.numeric(year))
```

Merge with shooting data:

```{r}
shooting_holidays <- full_join(shooting, holidays, by = c("month", "day", "year")) %>% 
  mutate(
    shooting = incident_key > 0,
    shooting = case_match(shooting,
                          TRUE ~ "Incident",
                          NA ~ "No Incident"),
    holiday_yn = is.na(holiday),
    holiday_yn = case_match(holiday_yn,
                            TRUE ~ "No Holiday",
                            FALSE ~ "Holiday")
  )
```

Plot of yes vs no holiday:

- [ ] make plot look better

```{r}
plot_holiday_yn <- shooting_holidays %>% 
  ggplot(aes(x = holiday_yn, y = shooting)) +
  geom_bar(stat = "identity", width = 0.5) +
  theme_minimal()

plot_holiday_yn
```

Plot of individual holidays vs no holiday:

- [ ] make plot look better

```{r}
plot_holiday_i <- shooting_holidays %>% 
  ggplot(aes(x = holiday, y = shooting)) +
  geom_bar(stat = "identity", width = 0.7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("")

plot_holiday_i
```

It seems that there's a trend of more shootings happening on non-holiday days than on holiday days. Will need to do further testing to confirm whether this is a significant association.

- [ ] test the significance of this, ttest for yes/no holiday, unsure for other holidays. linear regression, probably?



### COVID Analysis

Plot of per-year shooting incidence:

```{r}
plot_covid_yearly <- shooting %>% 
  ggplot(aes(x = year)) +
  geom_histogram(bins = 19, color = "violetred4", fill = "violetred4") +
  theme_minimal() +
  ylab("Number of Shootings") +
  xlab("Year")

plot_covid_yearly
```

There appears to be a decrease leading to 2020, and then a spike at 2020.

Create variable for whether or not shooting takes place during COVID-19 pandemic :

```{r}

```

Plot of yes/no COVID-19 shooting (NOT counts):

```{r}

```



### Time of Day Analysis

Create variable for day/night:

```{r}

```

Plot of day/night shootings:

```{r}

```

Plot of shootings per hour of the day:

```{r}

```



### Day of Week Analysis