---
title: "Murder flag"
author: "Miho Kawanami"
date: "2025-11-11"
output: html_document
---

# Introduction
This analysis aims to understand the distribution and determinants of fatal shooting incidents in New York City. Using the NYPD Shooting Incident data (historic and 2025), I describe borough-level patterns of murder occurrences and conduct a logistic regression model to identify individual characteristics associated with the likelihood that a shooting results in death (Murder Flag = 1).

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(tidycensus)
library(sf)
library(mapview)
```

```{r}
census_api_key("71edc420ec0b86cf5a54c3fc72727f4149467e79")
```

```{r}
library(tidyverse)
dat1 = read.csv("./Data Folder/Shooting_Historic.csv")
dat2 = read.csv("./Data Folder/Shooting_2025.csv")

dat1 = dat1 |>
  mutate(STATISTICAL_MURDER_FLAG = STATISTICAL_MURDER_FLAG %in% c(TRUE, "true", "True"))

dat2 = dat2 |>
  mutate(STATISTICAL_MURDER_FLAG = STATISTICAL_MURDER_FLAG %in% c(TRUE, "true", "True"))

dat_all = bind_rows(
dat1 |> mutate(source = "historic"),
dat2 |> mutate(source = "y2025")
)
```

## By Each Borough_historic
```{r}
dat1_boro = dat1 |>
  group_by(BORO) |>
  summarise(
    n_murder = sum(STATISTICAL_MURDER_FLAG),  
    n_total = n(),                             
    prop_murder = n_murder / n_total,
    .groups = "drop"
  )

dat1_boro
```

## By Each Borough_all
```{r boro-prop, message=FALSE, warning=FALSE}
dat_all_boro_prop = dat_all |>
  group_by(BORO) |>
  summarise(
    n_murder = sum(STATISTICAL_MURDER_FLAG),
    n_total  = n(),
    prop_murder = n_murder / n_total,
    .groups = "drop"
  )

dat_all_boro_prop
```

```{r boro-count, message=FALSE, warning=FALSE}
dat_all_boro_count = dat_all |>
  group_by(BORO) |>
  summarise(
    n_murder = sum(STATISTICAL_MURDER_FLAG),
    n_total  = n(),
    .groups = "drop"
  )

dat_all_boro_count
```

```{r boro-map-data, message=FALSE, warning=FALSE}
nyc_boro = tigris::counties(state = "NY", cb = TRUE) |>
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

boro_to_county = function(df) {
  df |>
    mutate(
      county_name = case_when(
        BORO == "MANHATTAN"     ~ "New York",
        BORO == "BROOKLYN"      ~ "Kings",
        BORO == "BRONX"         ~ "Bronx",
        BORO == "QUEENS"        ~ "Queens",
        BORO == "STATEN ISLAND" ~ "Richmond"
      )
    )
}

dat_all_boro_prop = boro_to_county(dat_all_boro_prop)
dat_all_boro_count = boro_to_county(dat_all_boro_count)

```

## Map_ percentage__historic

```{r message=FALSE, warning=FALSE}
nyc_boro = tigris::counties(state = "NY", cb = TRUE) |>
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

dat1_boro = dat1_boro |>
  mutate(
    county_name = case_when(
      BORO == "MANHATTAN" ~ "New York",
      BORO == "BROOKLYN" ~ "Kings",
      BORO == "BRONX" ~ "Bronx",
      BORO == "QUEENS" ~ "Queens",
      BORO == "STATEN ISLAND" ~ "Richmond"
    )
  )

nyc_boro_map = nyc_boro |>
  left_join(dat1_boro, by = c("NAME" = "county_name"))

mapview(nyc_boro_map,
        zcol = "prop_murder",
        layer.name = "Proportion of Murder by Borough",
        popup = paste0("Borough: ", nyc_boro_map$BORO,
                       "<br>Murder proportion: ",
                       round(nyc_boro_map$prop_murder, 3)))
```

## map (percentage)_all
```{r map-percentage, message=FALSE, warning=FALSE}
nyc_boro_map_prop = nyc_boro |>
  left_join(dat_all_boro_prop, by = c("NAME" = "county_name"))

mapview(
  nyc_boro_map_prop,
  zcol = "prop_murder",
  layer.name = "Proportion of Murder by Borough (all years)",
  popup = paste0(
    "Borough: ", nyc_boro_map_prop$BORO, "<br>",
    "Murder proportion: ", round(nyc_boro_map_prop$prop_murder, 3), "<br>",
    "Total shootings: ", nyc_boro_map_prop$n_total
  )
)
```

## map (count)_historic

```{r}
dat1_boro = dat1 |>
  group_by(BORO) |>
  summarise(
    n_murder = sum(STATISTICAL_MURDER_FLAG),
    n_total  = n(),
    .groups = "drop"
  )


dat1_boro <- dat1_boro |>
  mutate(
    COUNTY = case_when(
      BORO == "MANHATTAN" ~ "New York",
      BORO == "BROOKLYN" ~ "Kings",
      BORO == "BRONX" ~ "Bronx",
      BORO == "QUEENS" ~ "Queens",
      BORO == "STATEN ISLAND" ~ "Richmond"
    )
  )

library(tigris)
options(tigris_progress = FALSE, tigris_use_cache = TRUE)

nyc_boro_map <- counties(state = "NY", cb = TRUE) |>
  filter(NAME %in% c("Bronx", "Kings", "New York", "Queens", "Richmond"))

nyc_boro_map <- nyc_boro_map |>
  left_join(dat1_boro, by = c("NAME" = "COUNTY"))

mapview(
  nyc_boro_map,
  zcol = "n_murder",
  layer.name = "Number of Murder Cases by Borough",
  popup = paste0(
    "Borough: ", nyc_boro_map$BORO, "<br>",
    "Number of murder cases: ", nyc_boro_map$n_murder, "<br>",
    "Total shootings: ", nyc_boro_map$n_total
  )
)
```

## map (count)_all
```{r map-count, message=FALSE, warning=FALSE}
nyc_boro_map_count = nyc_boro |>
  left_join(dat_all_boro_count, by = c("NAME" = "county_name"))

mapview(
  nyc_boro_map_count,
  zcol = "n_murder",
  layer.name = "Number of Murder Cases by Borough (all years)",
  popup = paste0(
    "Borough: ", nyc_boro_map_count$BORO, "<br>",
    "Number of murder cases: ", nyc_boro_map_count$n_murder, "<br>",
    "Total shootings: ", nyc_boro_map_count$n_total
  )
)
```

## regression model 
```{r regression, message=FALSE, warning=FALSE}
dat_all_stable = dat_all |>
  filter(VIC_SEX %in% c("M","F")) |>                 
  filter(!VIC_AGE_GROUP %in% c("UNKNOWN", "", NA)) |>
  mutate(
    age_simple = case_when(
      VIC_AGE_GROUP %in% c("<18", "18-24") ~ "young",
      VIC_AGE_GROUP %in% c("25-44", "45-64") ~ "adult",
      VIC_AGE_GROUP == "65+" ~ "old"
    ),
    age_simple = factor(age_simple, levels = c("adult", "young", "old")),
    STATISTICAL_MURDER_FLAG = as.numeric(STATISTICAL_MURDER_FLAG),
    VIC_SEX = factor(VIC_SEX)
  )

fit_stable = glm(
  STATISTICAL_MURDER_FLAG ~ VIC_SEX + age_simple,
  data = dat_all_stable,
  family = binomial()
)

summary(fit_stable)

```

## OR
```{r}
library(broom)
tidy(fit_stable, exponentiate = TRUE, conf.int = TRUE)
```

* **Comments**
The regression results indicate that age is a significant predictor of fatal outcomes:
younger victims (under 25) are significantly less likely to die from a shooting compared to adults, while older victims (65+) have a higher likelihood of fatality. Sex was not significantly associated with the murder outcome. These findings are consistent with known patterns in trauma epidemiology, where older individuals face higher mortality risks from penetrating injuries.


